Experiences in Modelling a Microcontroller Instruction Set Using B

### 2012-01-09 trunk/papers/ArticleZ80/source/paper.tex
Change the related works (I think that it can be better).

  This paper describes an approach to model the functional aspects of
  the instruction set of microcontroller platforms and an embedded software
  verified up to assembly level. These models were developed using the notation
  of the B method. The paper presents specifically the case of the Z80 
  platform and case study software that can be used in tanks of the petroleum
  industry. This work is a contribution towards the extension of the B method
  to handle developments up to the assembly level code.
\end{abstract}
%
%\keywords{ Validation, Verification, Microcontroller, Hardware}
\paragraph[Keywords]{ \textbf{Keywords}: Validation, Verification,  Embedded Software and
Hardware}

@@

The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantee its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined up to the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} has presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

This work gives an overview of the formal modelling of the instruction
set of the Z80 microcontroller~\cite{Z80_manual}\footnote{The
  interested reader in more details is invited to visit our repository
  at: http://code.google.com/p/b2asm.}. Using the responsibility
division mechanism provided by B, auxiliary libraries of basic modules
were developed as part of the construction of microcontroller
model. Such library has many definitions about common concepts used in
the microcontrollers; besides the Z80 model, it is used by two other
microcontrollers models that are under way.

Other possible uses of a formal model of a microcontroller
instruction set include documentation, the construction of simulators,
and can be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover, the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine.

@@

In general, the manuals of microcontrollers can be more suited for
developers in assembly language. Many times, each instruction of microcontrollers
is shown in its official manual using textual description, math partially
description of instruction, examples, encoding, number of cycles and other
information. But the explanations about the action of instruction do not use
usually the best way to describe the instruction for users and developers. For
example, if the users want to know the action of instruction and its effects on
the flag register then he needs to search this information in different pages of
the manual. Besides, the writer of the manual usually describes important facts
in not formal notation and default. So, this description can add some
ambiguities or even mistakes.

% b) The case of Z80
The official manual of the microcontroller Z80 \cite{Z80_manual} has several
problems that were identified over time~\cite{UndocumentedZ80}. The Z80 is a
microcontroller that  has been tested and used for many years. This intensive use
and test facilitated to find several errors in the description of the
instructions in its manual. Several of these problems have been reported in a
technical report \cite{UndocumentedZ80}. Some problems are inaccurated
information, for example the description of action of instruction on the flag
register, partial information and distributed on different pages of manual.

A simple example of error in Z80 Manual happens
% \begin{itemize} \item Some input and output instructionsThe
% INI/INIR/IND/INDR/OUTI/OUTD/OTIR/OTDR instructions do affect the CF flag (some
% official documentation says they leave it unaffected, important!) and the NF
% flag is not always set but may also be reset (see 4.3 for exact operation)
% \cite{UndocumentedZ80}. Este erro he descrito na phegina (274 do documento e
% 294 do pdf) do manual oficial do Z80 (Z80_CPU_adotado.pdf) Citar apenas esse
% exemplo abaixo! \item
when a determined interruption (NMI) is accepted, the bit IFF1 is not
copied to bit IFF2~\cite{UndocumentedZ80}.  This example can cause serious
problems for programmers.
% Este erro he descrito na phegina (22 do documento e 26do pdf) do manual
% official do Z80 (Z80_CPU_adotado.pdf) Only IFF1 is reset.
This error was identified, because the description of interruption on the
official manual is big and complex, so to help the construction of Z80 B model,
other references \cite{UndocumentedZ80,Simulator_z80}  were also consulted.

An interesting solution is to specify formally the assembly instruction set, this
creates a pattern of representation and validates properties of instructions. The
formal model also restricts the definitions to correct typing and uses only
expressions well-defined.Furthermore, the developer can use with B method
different levels of abstraction and add more specific details in refined models.

Finally, the main goal of this project, though, is to provide a basis for the
generation of software artefacts at the assembly level, those are amenable to 
refinement verification within the B method.

\subparagraph{Report structure}

@@

The report is structured as it follows. Section~\ref{sec:B_method} provides a
short introduction to the B method. Section~\ref{sec:models} presents the elementary libraries and
the modelling of some elements common to microcontrollers.  Section~\ref{sec:z80}
presents the B model of the Z80 instruction set. Section ~\ref{sec:Verification}
provides some information on the proof effort needed to analyse the presented
models. Section ~\ref{sec:studycase} explains the building and verifying
assembly-level refinements. Related works are discussed in
section~\ref{sec:relatedworks}. Finally, the last section is devoted to the conclusions.

\section{Introduction to the B Method}
\label{sec:B_method}

The B method for software development~\cite{Abrial} is based on the B Abstract
Machine Notation (AMN) and the use of formally proved refinements up to a
specification sufficiently concrete that programming code can be automatically
generated from it. Its mathematical basis consists of first order logic, integer
arithmetic and set theory, and its corresponding constructs are similar to those
of the Z notation.

A B specification is structured in modules. A module defines a set of valid
states, including a set of initial states, and operations that may provoke a
transition among states. The design process starts with a module with a
so-called functional model of the system under development. In this initial
modelling stage, the B method requires that the user proves that, in a machine,
all its initial states are valid, and operations do not define
transitions from valid states to invalid states.

Essentially, a B module contains two main parts: a header and the available
operations. Figure~\ref{fig:maqB} has a very basic example. The clause
$\mathit{MACHINE}$ has the name of module.  The next two clauses, respectively,
reference external modules and create an instance of an external module. The
$\mathit{VARIABLES}$ clause declares the name of the variables that compose the
state of the machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $ specifies
the initial states. Finally, operations correspond to the transitions among
states of the machine.

\begin{figure}[h]
  % \begin{small}
  $$
  \begin{array}{lcl}
    \begin{array}[t]{l}
      \MACHINE \quad  \mathit{micro} \\
      \SEES \quad \mathit{TYPES}, \mathit{ALU} \\
      \INCLUDES  \quad \mathit{MEMORY} \\
      \VARIABLES \quad    \mathit{pc}  \\
      \INVARIANT\\ 
      \quad \mathit{pc} \in \mathit{INSTRUCTION} 
    \end{array}
    & \hspace*{0cm} &
    \begin{array}[t]{l}
      \INITIALISATION  \mathit{pc} := 0 \\
      \OPERATIONS\\
      \mathit{JMP} \mathit{( jump )} = \\
      \quad \PRE \mathit{jump} \in \mathit{INSTRUCTION}\\ 
      \quad \THEN \mathit{pc} := \mathit{jump}\\  
      \quad \END \\
      \END
    \end{array}
  \end{array}
  $$
\caption{A very basic B machine.}
\label{fig:maqB}
\end{figure}

@@

We have been developed a reusable set of basic definitions to model hardware concepts and data types
concepts. These definitions are grouped into three separate development projects
and are available as libraries. A fourth project is devoted to the higher-level aspects of the platform. Thus the workspace is
composed of: a power library\footnote{The power library has the basics definitions to help the theorems
prover about simple calculus of power, especially power of two. It has constants, that represents all
results of power function needed, and the definition of power function.}, a hardware library, a types
library and a project for the specific platform, in this case the Z80. The corresponding dependency
diagram is depicted in Figure~\ref{fig:hardware-definition-graph}; information specific to each project is
presented in the following.

\begin{figure}[h]
\centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.png}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Bit representation and manipulation}

The entities defined in the module $\mathit{BIT\_DEFINITION}$ are the
type for bits, logical operations on bits (negation, conjunction,
disjunction, exclusive disjunction), as well as a conversion function
from Boolean to bit.

First, bits are modelled as a set of integers: $\mathit{BIT} =
\mathit{0..1}$. The negation is an unary function on bits and it is
defined as:

$
\begin{array}{l}
\mathit{bit\_not}  \in  \mathit{BIT}  \fun  \mathit{BIT}  \land \forall ( \mathit{bb}). (\mathit{bb} \in \mathit{BIT} \implies \mathit{bit\_not}(\mathit{bb}) =
1-\mathit{bb})\\
\end{array}
$

@@

$
\begin{array}{l}
%  \mathit{bit\_not}(0) = 1;  \mathit{bit\_not}(1) = 0; \\
\forall (\mathit{bb}).(\mathit{bb} \in \mathit{BIT} \implies \mathit{bit\_not}(\mathit{bit\_not}(\mathit{bb})) = \mathit{bb})
\end{array}
$

Conjunction is a unary function on bits and it is defined as:

$
\begin{array}{l}
\mathit{bit\_and} \in \mathit{BIT} \times \mathit{BIT} \fun \mathit{BIT} \land \\
\forall (\mathit{b1}, \mathit{b2}).(\mathit{b1}  \in \mathit{BIT}  \land \mathit{b2} \in \mathit{BIT} \implies \\
\quad ((\mathit{bit\_and}(\mathit{b1}, \mathit{b2}) = 1) \iff (\mathit{b1} = 1)  \land  (\mathit{b2} = 1)))
\end{array}
$

The module provides the following lemmas for conjunction, either:

$
\begin{array}{l}
%  \mathit{bit\_and}(0,0) = 0;  \mathit{bit\_and}(0,1) = 0; \\
%  \mathit{bit\_and}(1,0) = 0;  \mathit{bit\_and}(1,1) = 1; \\
\forall (\mathit{b1},\mathit{b2}).(\mathit{b1} \in \mathit{BIT} \land \mathit{b2} \in \mathit{BIT} \implies \\
\quad (\mathit{bit\_and}(\mathit{b1}, \mathit{b2}) = \mathit{bit\_and}(\mathit{b2},\mathit{b1})))\land \\
\forall (\mathit{b1},\mathit{b2},\mathit{b3}).(\mathit{b1} \in \mathit{BIT} \land  \mathit{b2} \in \mathit{BIT} \land \mathit{b3} \in \mathit{BIT} \implies \\
\quad (\mathit{bit\_and}(\mathit{b1}, \mathit{bit\_and}(\mathit{b2},\mathit{b3})) = \mathit{bit\_and}(\mathit{bit\_and}(\mathit{b1},\mathit{b2}),\mathit{b3})))\\
% \forall (\mathit{b1}).(\mathit{b1} \in \mathit{BIT} \implies (\mathit{bit\_and}(\mathit{b1}, 1) = \mathit{b1})); \\
% \forall (\mathit{b1}).(\mathit{b1} \in \mathit{BIT} \implies (\mathit{bit\_and}(\mathit{b1}, 0) = 0));
\end{array}
$

The module provides definitions of $\mathit{bit\_or}$ (disjunction)
and $\mathit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their expression in B is
similar as for $\mathit{bit\_and}$, they are thus omitted.

@@

$
\begin{array}{l}
\mathit{bool\_to\_bit} \in \BOOL \fun \mathit{BIT} \land \mathit{bool\_to\_bit} = \{ \TRUE \mapsto 1, \FALSE \mapsto 0 \} \\
\end{array}
$

@@

\subsection{Modelling bytes and bit vectors of length 16}

Bit vectors of length 8 are bytes. They form a common entity in
hardware design. We provide the following definitions:

\hspace*{0.0in}\it BYTE\_WIDTH \rm = 8 $\land$ \it BYTE\_INDEX \rm = 1 $\upto$ \rm  BYTE\_WIDTH\rm $\land$

@@

\hspace*{0.0in}\it BYTE \rm = \rm \{ \it bt  $\mid$  \it bt $\in$ \it BIT\_VECTOR  $\land$  \it bv\_size\rm (\it bt\rm )\rm =\it BYTE\_WIDTH\rm \}\hspace*{0.10in} $\land$

\hspace*{0.0in}\it BYTE\_ZERO  $\in$  \it BYTE  $\land$ \it BYTE\_ZERO \rm = \it BYTE\_INDEX  $\times$  \rm \{\rm 0\rm \}

The $\mathit{BYTE\_INDEX}$ is the domain of the functions modelling bytes. It starts at 1 to obey a
definition of sequences from B. However, it is common in hardware architectures to start indexing from
zero. The definition $\mathit{PHYS\_BYTE\_INDEX}$ is used to provide functionalities obeying this
convention. The $\mathit{BYTE}$ type is a specialized type from $\mathit{BIT\_VECTOR}$, but it has a size
limit. Other specific definitions are provided to facilitate further modelling: the type $\mathit{BV16}$
is created for bit vector of length 16 in a similar way.

\subsection{Bit vector arithmetics}

Bit vectors are used to represent and combine numbers: integer ranges (signed or unsigned). Therefore, our
library includes functions to manipulate such data, for example, the function $\mathit{bv\_to\_nat}$ that
maps bit vectors to natural numbers:

$
\begin{array}{l}
\mathit{bv\_to\_nat} \in \mathit{BIT\_VECTOR} \fun \nat \land \\
\mathit{bv\_to\_nat} = \lambda v . (v \in \mathit{BIT\_VECTOR} \mid \sum i . (i \in \dom(v) . v(i)
\times 2^{i-1}))
\end{array}
$

An  associated lemma is: $\forall n . (n \in \nat_1 \implies \mathit{bv\_to\_nat}(\mathit{nat\_to\_bv}(n)) = n)$

@@

The instruction set of microcontrollers usually have common data types. These types are placed in the
types library. Each type module has functions to manipulate and convert its data. There are six common
basics data types represented by modules, see details in table~\ref{tab:types}.

\begin{table}
\caption{Descriptions of basic data types}
\label{tab:types}

% \begin{center}
% \begin{tabular}{|c|c|c|c|c|c|c|}
% \hline
%  $Type\ Name$ & UCHAR & SCHAR & USHORTINT & SSHORTINT  & BYTE & BV16 \\\hline
%  $Range$ & 0..255 & -128..127 & 0..65.535  & -32.768..32.767  & -- & --\\ \hline
%  $Physical\ Size $ & 1 byte & 1 byte & 2 bytes & 2 bytes &  1 bytes & 2 bytes \\ \hline
% \end{tabular}
% \end{center}

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
 $Type\ Name$ & $Range$ & $Physical\ Size $\\\hline
 UCHAR & 0..255 & 1 byte\\\hline
 SCHAR & -128..127 & 1 byte  \\\hline
 USHORTINT & 0..65.535 & 2 byte \\\hline
 SSHORTINT & -32.768..32.767 & 2 byte \\\hline
 BYTE & -- & 1 bytes  \\\hline
 BV16 & -- & 2 bytes \\ \hline
\end{tabular}
\end{center}

\end{table}

Usually, each type module just needs to instantiate concepts that were already defined in the hardware
modelling library.  For example, the function $\mathit{bv\_to\_nat}$ from bit vector arithmetic is
specialized to $\mathit{byte\_uchar}$. As the set $\mathit{BYTE}$ is a subset of the
$\mathit{BIT\_VECTOR}$, this function can defined as follows:

$
\begin{array}{l}
\mathit{byte\_uchar} \in \mathit{BYTE} \fun \nat \land \\
\mathit{byte\_uchar} = \lambda (v) . ( v \in BYTE | bv\_to\_nat(v) )
\end{array}
$

@@

The inverse function $\mathit{uchar\_byte}$ is easily defined:

$
\begin{array}{l}
\mathit{uchar\_byte} \in \mathit{UCHAR}  \fun  \mathit{BYTE}  \land \\
   \mathit{uchar\_byte} = \ (\mathit{byte\_uchar}) ^{-1}
\end{array}
$

>>>>

  This paper describes an approach to model the functional aspects of the instruction set of
  microcontroller platforms and several details about the representation of
  elements from microcontrollers.  Several models were developed using the notation of the B method. 
  They are used to develop a formally verified software up to the assembly level and allow the
  simulation of models. This simulation is able to guarantee the consistency between the execution
  of a software model and a real  execution of software, it has wide variety of usages in industry
  and academia. This paper presents specifically the case of the Z80 platform and quote a theoretic 
  case study important in tanks of the petroleum industry. This work is a contribution towards the
  extension of the B method to handle developments up to the assembly level code.

%  \paragraph[Keywords]{
   \textbf{Keywords}: Embedded Software, Simulation, B method and Verification%}
\end{abstract}

%\keywords{ Validation, Verification, Microcontroller, Hardware}
%
%\setcounter{secnumdepth}{1}
%\setcounter{chapter}{1}
%\setcounter{section}{1}

@@

The B method~\cite{Abrial} supports the construction of safety systems
models by generating proof obligation that must be verified to guarantee its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined up to the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. \cite{Dantas_SBMF08} proposed
an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

This work presents a formal modell of the instruction set of the Z80 
microcontroller~\cite{Z80_manual}. This microcontroller
was chosen because of several factors: it contains essential and common
concepts in microcontrollers and until microprocessors; it has extensive
documentation available; and it was widely used and is still in use. 
A first Z80 modell presented in \cite{Valerio_SBMF09} was verified completely,
but it uses constructs that are not well supported to animate. 
This modelling was changed to support animation in ProB\cite{proB}.
Basically, the new modelling\footnote{The interested reader in more
details is invited to visit our repository at: http://code.google.com/p/b2asm/.}
changed elements represented by infinite sets and adjusted the implications
of this change. Using the responsibility division mechanisms provided by B, auxiliary
libraries of basic modules were developed as part of the construction of microcontroller
model. Such library has many definitions about common concepts used in
the microcontrollers;besides the new Z80 model, a theoretic case study in 
petroleum production test system was developed using Z80 assembly language
to analyse the code simulation and verification process.

The formal model of instruction set provides many benefits. The animation of 
Z80 model allows to simulate the execution assembly programs including
support for interrupts and input and output instructions. Other possible uses
of a formal model of a microcontroller instruction set include documentation,
the construction of simulators, and can be possibly the starting point of a 
verification effort for the actual implementation of a Z80 design. Moreover, 
the model of the instruction set could be instrumented with non-functional
aspects, such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine.
Two aspects of these formalization are of particular importance to us:
provide a more solid documentation artifact for microcontrollers, 
build a reference model for a formal compilation approach in B.

@@

In general, the manuals of microcontrollers can be more suited for
developers in assembly language. Many times, each instruction of microcontrollers
is shown in its official manual using textual description, math description, examples,
encoding, number of cycles and other information.Some times, the instruction descriptions
have an own notation, some informations are organized in different sections of document 
and the textual description are not standardized. This does not allow the interpretation by parser.
% b) The case of Z80
For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users
want to know the action of an instruction and its effects on the flag register then he needs
to read an informal textual description. So, this description can add
some ambiguities or even mistakes. The official manual has also several problems that were 
identified over time~\cite{UndocumentedZ80}. The Z80 is a microcontroller that  has been
tested and used for many years. This intensive use and test facilitated to find several
errors in the description of the instructions in its manual. Several of these problems
have been reported in a technical report \cite{UndocumentedZ80}. Some problems are inaccurate
information, for example the description of the action of an instruction on the flag
register, partial information and distributed on different pages of manual.

% in the Z80 manual happens was that,

%A simple example of error is in some official documentation saying that a group of instructions leaves the \textit{CF} flag unaffected and the \textit{NF}
%flag is not always set, but may also be reset \cite{UndocumentedZ80}.
%% \begin{itemize} \item Some input and output instructionsThe
%% INI/INIR/IND/INDR/OUTI/OUTD/OTIR/OTDR instructions do affect the CF flag (some
%% official documentation says they leave it unaffected, important!) and the NF
%% flag is not always set but may also be reset (see 4.3 for exact operation)
%% \cite{UndocumentedZ80}. Este erro he descrito na phegina (274 do documento e
%% 294 do pdf) do manual oficial do Z80 (Z80_CPU_adotado.pdf) Citar apenas esse
%% exemplo abaixo! \item
%%when a determined interruption (NMI) is accepted, the bit IFF1 is not   
%%copied to bit IFF2~\cite{UndocumentedZ80}. 
%%
%This example can cause serious problems for programmers.
%% Este erro he descrito na pagina (22 do documento e 26do pdf) do manual
%% official do Z80 (Z80_CPU_adotado.pdf) Only IFF1 is reset.
%This error and others 7 different errors were previously identified by~\cite{UndocumentedZ80}. So to help the 
%construction of Z80 B model, other references~\cite{UndocumentedZ80,Simulator_z80}
%were also consulted.

A solution to avoid ambiguity and inconsistencies is to specify formally the
assembly instruction set, this creates a pattern of representation and validates
properties of instructions. The formal model also restricts the definitions to
correct typing and uses only well-defined expressions. Furthermore, the developer
can use with B method different levels of abstraction and add more specific details
in refined models.

%Finally, the main goal of this project, though, is to provide a basis for the
%generation of software artefacts at the assembly level, those are amenable to 
%refinement verification within the B method.

%\subparagraph{Report structure}

@@

This paper is structured as it follows. %Section~\ref{sec:B_method} provides a short introduction to the B method. 
Section~\ref{sec:models} presents the elementary libraries and
the modelling of some elements common to microcontrollers.  Section~\ref{sec:z80}
presents the B model of the Z80 instruction set. %Section ~\ref{sec:Verification} provides some information on the proof effort needed to analyse the presented models.
Section ~\ref{sec:studycase} explains the building and verifying assembly-level refinements. Related works are discussed in
section~\ref{sec:relatedworks}. Finally, the last section is devoted to the conclusions.
%
%\section{Introduction to the B Method}
%\label{sec:B_method}
%
%The B method for software development~\cite{Abrial} is based on the B Abstract
%Machine Notation (AMN) for the formal specification of functional requirements and the use of formally proved refinements up to a
%specification sufficiently concrete that programming code can be automatically
%generated from it. Its mathematical basis consists of first order logic, integer
%arithmetic and set theory, and its corresponding constructs are similar to those
%of the Z notation~\cite{Spivey96}.
%
%A B specification is structured in modules. A module defines a set of valid
%states, including a set of initial states, and operations that may provoke a
%transition among states. The design process starts with a module with a
%so-called functional model of the system under development. In this initial
%modelling stage, the B method requires that the user proves that, in a machine,
%all its initial states are valid, and operations do not define
%transitions from valid states to invalid states.
%
%
%Essentially, a B module contains two main parts: a header and the available
%operations. Figure~\ref{fig:maqB} has a very basic example. The clause
%\MACHINE  has the name of module.  The next two clauses, respectively,
%reference external modules and create an instance of an external module.  The
%\VARIABLES clause declares the name of the variables that compose the
%state of the machine. Next, the \INVARIANT clause defines the valid states,
%specifying  the type and other restrictions on the variables. The \INITIALISATION specifies
%the initial states. Finally, operations correspond to the transitions among
%states of the machine.
%
%
%\begin{figure}[h]
%  % \begin{small}
%\begin{center}
%\begin{sloppypar}
%
%  $\begin{array}{lcl}
%    \begin{array}[t]{l}
%      \MACHINE \quad  \textit{micro} \\
%      \SEES \quad \textit{TYPES}, \textit{ALU} \\
%      \INCLUDES  \quad \textit{MEMORY} \\
%      \VARIABLES \quad    \textit{pc}  \\
%      \INVARIANT\\ 
%      \quad \textit{pc} \in \textit{INSTRUCTION} 
%    \end{array}
%    & \hspace*{0cm} &
%    \begin{array}[t]{l}
%      \INITIALISATION  \textit{pc} := 0 \\
%      \OPERATIONS\\
%      \textit{JMP} \textit{( jump )} = \\
%      \quad \PRE \textit{jump} \in \textit{INSTRUCTION}\\ 
%      \quad \THEN \textit{pc} := \textit{jump}\\  
%      \quad \END \\
%      \END
%    \end{array}
%  \end{array}
%  $
% \end{sloppypar} 
%\end{center}
%\caption{A very basic B machine.}
%\label{fig:maqB}
%\end{figure}

@@

We have developed a reusable set of basic definitions to model hardware concepts and data types
concepts. The models contains definitions to represent the: registers, interruptions, input and output
ports, memory and instructions. These definitions are grouped into separated development projects
and are available as libraries.

Thus the workspace is composed of modules group and libraries: 
\begin{itemize}

\item \textbf{Power}: it has the basics definitions to help the theorems
prover about simple calculus of power, especially power of two. It contains the definition of power function
and it has constants, that represent all required results of power function.

\item \textbf{Hardware Library}: it has the definition of bit vectors with size 8 and 16, basic functions to manipulate bit vectors and important assertions. This is presented in sections \ref{subsec:HardwareLibrary1}, \ref{subsec:HardwareLibrary2} and \ref{subsec:HardwareLibrary3}.

 \item \textbf{Types Library}: it has the definition of naturals and integers represented with  8 bits and 16 bits, basic conversion functions between the types and bit vectors and important assertions. This is presented in section \ref{subsec:Types}.

 \item \textbf{Platform}: it has the definition functions of arithmetic logic unit,  memory unit, registers and assembly instructions. This is presented in section \ref{sec:z80}.

 The corresponding dependency
diagram is depicted in Figure~\ref{fig:hardware-definition-graph}; information specific to each project is
presented in the following.

\begin{figure}[h]
\centering
\includegraphics[width=.77\textwidth]{diagramaEstrutural_vertical_ProB.png}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Bit representation and manipulation}
\label{subsec:HardwareLibrary1}

The entities defined in the module $\textit{BIT\_DEFINITION}$ are the
type for bits, logical operations on bits (negation, conjunction,
disjunction, exclusive disjunction), as well as a conversion function
from Boolean to bit.

First, bits are modelled as a set of integers: $\textit{BIT} =
\textit{0..1}$. The negation is an unary function on bits and it is
defined as:

$
\begin{array}{l}
\textit{bit\_not}  \in  \textit{BIT}  \fun  \textit{BIT}  \land 
\textit{bit\_not} \rm = \rm \{ \rm 0  \mapsto  \rm 1 , \rm 1  \mapsto  \rm 0 \rm \}
\end{array}
$
%\forall ( \textit{bb}). (\textit{bb} \in \textit{BIT} \implies {bit\_not}(\textit{bb}) =1-\textit{bb})\\

@@

$
\begin{array}{l}
%  \textit{bit\_not}(0) = 1;  \textit{bit\_not}(1) = 0; \\
\forall (\textit{bb}).(\textit{bb} \in \textit{BIT} \implies \textit{bit\_not}(\textit{bit\_not}(\textit{bb})) = \textit{bb})
\end{array}
$

Conjunction is a binary function on bits and it is defined as:

$
\begin{array}{l}
\textit{bit\_and} \in \textit{BIT} \times \textit{BIT} \fun \textit{BIT} \land \\
\forall (\textit{b1}, \textit{b2}).(\textit{b1}  \in \textit{BIT}  \land \textit{b2} \in \textit{BIT} \implies \\
\quad ((\textit{bit\_and}(\textit{b1}, \textit{b2}) = 1) \iff (\textit{b1} = 1)  \land  (\textit{b2} = 1)))
\end{array}
$

The module provides the following lemmas for commutativity and associativity:

$
\begin{array}{l}
%  \textit{bit\_and}(0,0) = 0;  \textit{bit\_and}(0,1) = 0; \\
%  \textit{bit\_and}(1,0) = 0;  \textit{bit\_and}(1,1) = 1; \\
\forall (\textit{b1},\textit{b2}).(\textit{b1} \in \textit{BIT} \land \textit{b2} \in \textit{BIT} \implies \\
\quad (\textit{bit\_and}(\textit{b1}, \textit{b2}) = \textit{bit\_and}(\textit{b2},\textit{b1})))\land \\
\forall (\textit{b1},\textit{b2},\textit{b3}).(\textit{b1} \in \textit{BIT} \land  \textit{b2} \in \textit{BIT} \land \textit{b3} \in \textit{BIT} \implies \\
\quad (\textit{bit\_and}(\textit{b1}, \textit{bit\_and}(\textit{b2},\textit{b3})) = \textit{bit\_and}(\textit{bit\_and}(\textit{b1},\textit{b2}),\textit{b3})))\\
% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 1) = \textit{b1})); \\
% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 0) = 0));
\end{array}
$

The module provides definitions of $\textit{bit\_or}$ (disjunction)
and $\textit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their expression in B is
similar as for $\textit{bit\_and}$, they are thus omitted from this paper.

@@

$
\begin{array}{l}
\textit{bool\_to\_bit} \in \BOOL \fun \textit{BIT} \land \textit{bool\_to\_bit} = \{ \TRUE \mapsto 1, \FALSE \mapsto 0 \} \\
\end{array}
$

@@

\subsection{Modelling bytes and bit vectors of length 16}
\label{subsec:HardwareLibrary3}

Bit vectors of length 8 are bytes. They are a common entity in
hardware design. We provide the following definitions:

\hspace*{0.0in}\it BYTE\_WIDTH \rm = 8 $\land$ \it BYTE\_INDEX \rm = 1 $\upto$ \rm  BYTE\_WIDTH\rm  \hspace*{0.05in} $\land$

@@

\hspace*{0.0in}\it BYTE \rm =\rm (\it BYTE\_INDEX  $\fun$  \it BIT\rm )\hspace*{0.10in} $\land$  \bf card\rm (\it BYTE\rm )\rm = $2^{8}$  $\land$

\hspace*{0.0in}$\forall$ \rm (\it b1\rm )\rm .\rm (\it b1  $\in$  \it BYTE  $\implies$  \bf size\rm (\it b1\rm )\rm =\it BYTE\_WIDTH  $\land$  \it b1  $\in$  \bf seq1\rm (\it BIT\rm )\rm )

%\hspace*{0.0in}\it BYTE \rm = \rm \{ \it bt  $\mid$  \it bt $\in$ \it BIT\_VECTOR  $\land$  \it size\rm (\it bt\rm )\rm =\it BYTE\_WIDTH\rm \} $\land$

%\hspace*{0.0in}\it BYTE\_ZERO  $\in$  \it BYTE  $\land$ \it BYTE\_ZERO \rm = \it BYTE\_INDEX  $\times$  \rm \{\rm 0\rm \}

The $\textit{BYTE\_INDEX}$ is the domain of the functions modelling bytes. It starts at 1 to obey a
definition of sequences from B. However, it is common in hardware architectures to start indexing from
zero. The definition $\textit{PHYS\_BYTE\_INDEX}$ is used to provide functionalities obeying this
convention. The $\textit{BYTE}$ type is as a specialized type of $\textit{BIT\_VECTOR}$, but it has a size
limit. Other specific definitions are provided to facilitate further modelling: the type $\textit{BV16}$
is created for bit vector of length 16 in a similar way.

\subsection{Bit vector arithmetic}
\label{subsec:Types}

Bit vectors are used to represent and combine numbers: integer ranges (signed or unsigned). Therefore, our
library includes functions to manipulate such data, for example, the function $\textit{bv\_to\_nat}$ that
maps bit vectors to natural numbers:

$
\begin{array}{l}
\textit{bv\_to\_nat} \in \textit{BIT\_VECTOR} \fun \nat  \land \\
\textit{bv\_to\_nat} = \lambda v .  (v \in \textit{BIT\_VECTOR} \mid  \sum i . (i \in \dom(v) . v(i)
\times 2^{i-1}))
\end{array}
$

An  associated lemma is: $\forall n . (n \in \nat_1 \implies \textit{bv\_to\_nat}(\textit{nat\_to\_bv}(n)) = n)$

@@

The instruction set architecture usually have common bit vector type and integer type.
In table~\ref{tab:types}, the first three are placed in the hardware library and the last four types are placed in the integer types library. 
Each type module has functions to manipulate and convert its data. There are seven common basic data types represented by modules.

\begin{table}
\caption{Details of basic data types}
\label{tab:types}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 $Type\ Name$ & $Range$ & $Physical\ Size $  & $Associated\ Proof\ Obligations$ \\\hline
 BIT & 0..1 & 1 bit &   118 \\\hline %49 + 69 = 118 \\\hline
 BYTE & -- & 1 bytes &  153\\\hline%17 + 136 = 153\\\hline
 BV16 & -- & 2 bytes &  75 \\ \hline%5 + 70 = 75 \\ \hline
 UCHAR & 0..255 & 1 byte &  30 \\\hline %4 + 26 = 30 \\\hline
 SCHAR & -128..127 & 1 byte & 26 \\\hline %2 + 24 = 26 \\\hline
 USHORT & 0..65.535 & 2 byte & 62 \\\hline % 4 + 58 = 62 \\\hline
 SSHORT & -32.768..32.767 & 2 byte & 58\\\hline %  2 + 56 = 58\\\hline
%  &  & Total & 522\\\hline %  2 + 56 = 58\\\hline
\end{tabular}
\end{center}
\end{table}

Usually, each type module just needs to instantiate concepts that were already defined in the hardware
modelling library.  For example, the function $\textit{bv\_to\_nat}$ from bit vector arithmetic is
specialized to $\textit{byte\_uchar}$. As the set $\textit{BYTE}$ is a subset of the
$\textit{BIT\_VECTOR}$, this function can defined as follows:

$
\begin{array}{l}
\textit{byte\_uchar} \in \textit{BYTE} \fun \nat \land \\
\textit{byte\_uchar} = \lambda (v) . ( v \in BYTE | bv\_to\_nat(v) )
\end{array}
$

@@

%The inverse function $\textit{uchar\_byte}$ is easily defined:
%
%$
%\begin{array}{l}
%\textit{uchar\_byte} \in \textit{UCHAR}  \fun  \textit{BYTE}  \land \\
%   \textit{uchar\_byte} = \ (\textit{byte\_uchar}) ^{-1}
%\end{array}
%$

 We also created the following lemmas:

 $
 \begin{array}{l}
  \forall (val) . (val \in \textit{UCHAR} |
  \textit{byte\_uchar}(\textit{uchar\_byte}(val)) = val) \land\\
  \forall (by) . (by \in \textit{BYTE} |
  \textit{uchar\_byte}(\textit{byte\_uchar}(by)) = by)
 \end{array}
 $

Similarly, several other general functions and lemmas were created for all other data types.
%[NEW PARAGRAPH-TO-REVIEW]
%Similarly, several other general functions and lemmas were created for all other data types.
%Some times, we need create arithmetic and logic functions that are more specific for a
%determined computer architecture. These functions are developed in the module ALU 
%(arithmetic  logic unit). The following shows some examples of specific functions used in
%microcontroller model of Z80. For example, the function $\textit{instruction_next}$
%receives a $\textit{BYTE}$ and returns a $\textit{BYTE}$ that contains a sequence of bit:
%a signal bit,

Nearly 50\% of all these proof obligations shown in the table~\ref{tab:types}
are verified almost automatically. Although many others proof obligations are difficult to verify, because these
check the type of functions with non-linear arithmetic is a difficult task.

\subsection{Verification Process in Hardware Library and Types Library} 
\label{sec:VerificationHardwareLibrary}

\input{VerificationLibraryHardware}
